---
title: "Dynamic Occupancy Modeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r data}
# Data from Kéry, M., Guillera‐Arroita, G., & Lahoz‐Monfort, J. J. (2013). 
# Analysing and mapping species range dynamics using occupancy models. 
# Journal of Biogeography, 40(8), 1463-1474.
data(crossbill, package = "unmarked")

# Get into format for HMM
nsites <- nrow(crossbill)
dat <- data.frame(ID = rep(1:nsites, each = 9),
                  year = rep(1:9, nsites), 
                  elev = rep(crossbill$ele, each = 9), 
                  forest = rep(crossbill$forest, each = 9), 
                  surveys = rep(crossbill$surveys, each = 9))
y <- apply(as.matrix(crossbill[,5:31]), 1, FUN = function(x) {
    tapply(as.numeric(x), rep(1:9, each = 3), FUN = function(r) {sum(r, na.rm = TRUE)})
})
y <- as.numeric(y)
dat$y2 <- ifelse(dat$surveys == 2, y, NA)
dat$y3 <- ifelse(dat$surveys == 3, y, NA)
dat$forest <- as.numeric(dat$forest)
dat$elev <- as.numeric(dat$elev)


head(dat)
```

The variables are:

- `ID`: identifier for site

- `year`: year of survey

- `surveys`: number of surveys that year (either 2 or 3)

- `y2`: number of observations in years with 2 surveys (0, 1, or 2)

- `y3`: number of observations in years with 3 surveys (0, 1, 2, or 3)

- `elev`: elevation in meters

- `forest`: forest cover 

The variables `y2` and `y3` can be analysed in a joint HMM, although they have to be included as two different data streams because they have different distributions. They are binomial with $n = 2$ or 3, and with parameter $p$ representing the detection probability. The 2-state underlying process represents occupancy (i.e., state 1 = occupied, state 2 = not occupied), and it is partially observed because we know that a site is occupied if $y > 0$. We also know that the detection probability is zero in state 2.

To simplify the process, we'll convert any value > 0 to 1, simply to denote detection. 

```{r}
dat$detect <- rep(NA, dim(dat)[1])
sr3 <- which(dat$surveys == 3)
sr2 <- which(dat$surveys == 2)

for(i in 1:dim(dat)[1]){
  if(i %in% sr2){
    if(dat$y2[i] == 0){
      dat$detect[i] <- 0}
    else{
      dat$detect[i] <- 1  
      }
  } else{
      if(dat$y3[i] == 0){
        dat$detect[i] <- 0
      }else{
        dat$detect[i] <- 1
      }
    }
  }

head(dat)
```

```{r}

library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

occ.data <- list(Tlen = dim(dat)[1], 
                 y = dat$detect, 
                 N = 2,
                 index = dat$ID)

occ.fit <- stan(file = "OccupancyEFI.stan", data = occ.data)

```

