---
title: "Simulating Data from a N-State HMM"
output:
 html_notebook
---

Let's simulate data from an HMM. As mentioned before, we need to specify a few things: 

- the number of states $N$
- the state-dependent distributions, $f(X_t | C_t = n)$
- the transition probabiility matrix
- the initial state distribution

We'll make use of the library `tidyverse`, which includes `ggplot2`, to plot our data. 

```{r, message=FALSE}
library(tidyverse)
```


```{r model_setup}
## Specify the number of states ---------------
N <- 2

## Specify the initial distribution -----------
init <- rep(1/N, N)

## Specify the transition probability matrix ----
tpm <- matrix(c(0.9, 0.1, 
                0.01, 0.99), 
              byrow=T, nrow=2)

## Specify the state-dependent distribution + values of parameters
## Normal(mean, standard deviation)
mu <- c(5, 10)
sd <- c(2, 1)

#pois_sdd <- c(4, 20)
```

Now let's simulate data from our 2-state HMM with normal state-dependent distributions. 

```{r simulate_data}
# Length of the time series
len <- 500

states <- numeric(len)
obs <- numeric(len)

# Simulating states
states[1] <- sample(size=1, x = 1:N, prob = init)

for(j in 2:len){
  states[j] <- sample(size=1, x = 1:N, prob = tpm[states[j-1],])
}

# Simulating observations
obs <- rnorm(len, mean=mu[states], sd = sd[states])
```

Visualize the simulated observations: 

```{r}
ggplot(data=tibble(x=1:len, y=obs, group=factor(states)), 
       aes(x, y, color=group)) + geom_point() + 
  xlab("Time") + ylab("Observations") + 
  theme_minimal() + scale_color_discrete(name="State")

```
Visualize the states: 

```{r}
ggplot(data=tibble(x=1:len, y=obs, group=factor(states)), 
       aes(x, y)) + geom_point(aes(color=group)) + 
  xlab("Time") + ylab("Observations") + 
  theme_minimal() + scale_color_discrete(name="States") + 
  geom_path(aes(x=1:len, y=obs), alpha=0.3)
```
```{r}

## plotting the state sequence

ggplot(data=tibble(x=1:len, states = states), aes(x, states)) + 
  geom_path() + theme_light() + xlab("Time") + ylab("States") 

```


State dwell-time distributions: 

```{r}
state1dur <- dgeom(x = 0:100, prob = 0.2)
state2dur <- dgeom(x=  0:100, prob = 0.3)

stateDur <- tibble(x=rep(1:101, 2),
                   dur = c(state1dur, state2dur), 
                   State=factor(rep(1:2, each=101)))

ggplot(stateDur, aes(x, dur, color=State)) + 
  geom_point(size=3) + theme_minimal() + 
  xlim(0, 25) + geom_path(size=2, alpha=0.4) + 
  ggtitle("State Dwell-Time Distributions") + 
  ylab("Probability") + xlab("Duration")
```

