
---
title: <center><font size="6"><b>R package overview</b></font></center>
subtitle: <center><font size="4"><b>Statistical Methods Seminar Series</b></font></center>
author: <center><font size="4"><a href="http://marceloarayasalas.weebly.com/">Marcelo Araya-Salas, PhD</a></font></center>
date: <center>`r format(Sys.Date(), "%d-%m-%Y")`</center>
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 3
    df_print: kable
    toc_float:
      collapsed: yes
      smooth_scroll: yes
fontsize: 12pt 
editor_options: 
  chunk_output_type: console
---

Installing packages:

```{r setup, include=FALSE, eval = FALSE}

remotes::install_github("maRce10/warbleR")
remotes::install_github("maRce10/ohun")
remotes::install_github("maRce10/PhenotypeSpace")
remotes::install_github("maRce10/baRulho")

```

Set working directory path:
```{r}

wd <- "./data"

```


# Rraven

Importing Raven selection tables into R:
```{r}

library(warbleR)
library(Rraven)

#load example data
data("selec.table")

# check txt files in ./data
list.files(path = wd, pattern = ".txt$")

# raven format
sels <- Rraven::imp_raven(path = wd)

# warbleR format
sels <- Rraven::imp_raven(path = wd, warbler.format = TRUE, all.data = FALSE)

# reading specific files
sels <- Rraven::imp_raven(path = wd, warbler.format = TRUE, all.data = FALSE, files = c("LBH 4 selection table example.txt", "LBH 2 selection table example.txt", "LBH 3 selection table example.txt"))

# extracting features 
## spectral parameters
warbleR::spectro_analysis(sels, path = wd)

# cross correlation
warbleR::cross_correlation(sels, path = wd)

# selection can also be exported to be open in Raven
Rraven::exp_raven(X = sels, path = wd, sound.file.path = wd, file.name = "single.file")

# and then re-exported into R
sngl <- Rraven::imp_raven(path = wd, files = "single.file.txt", warbler.format = TRUE, all.data = TRUE)


```


# Automatic detection: ohun



How to optimize detection
```{r}

data("lbh1", "lbh2", "lbh_reference")

# feature reference
ohun::feature_reference(lbh_reference)


# print spectrogram 1st file
ohun::label_spectro(wave = lbh1, reference = lbh_reference[lbh_reference$sound.files ==
    "lbh1.wav", ], hop.size = 10, ovlp = 50, flim = c(1, 10))

# 2nd file 
ohun::label_spectro(wave = lbh2, reference = lbh_reference[lbh_reference$sound.files ==
    "lbh2.wav", ], hop.size = 10, ovlp = 50, flim = c(1, 10))

# optimize detection
optim_detection <- ohun::optimize_energy_detector(reference = lbh_reference,
    threshold = c(2, 5, 10), min.duration = c(100), path = wd, smooth = c(5, 10, 20), bp = c(1.5, 9.6))
  
```

# once you have annotations: warbleR

formating and exploring sound files:
```{r}

# settting global parameters
warbleR::warbleR_options(path = wd)

# checking sound files
warbleR::check_sound_files()

# check format
warbleR::info_sound_files()

# consolidate
warbleR::consolidate()

# get durations  
warbleR::duration_sound_files()

```

quantify structure (feature extraction):
```{r}

# define global options
warbleR::warbleR_options(path = "./data", bp = "frange")

# measure spectral parameters
warbleR::spectro_analysis(X = lbh_selec_table, bp = c(0, 11), wl = 100, threshold = 1, ovlp = 10)

# measure mel frequency cepstral coefficients
warbleR::mfcc_stats(lbh_selec_table, numcep = 3)

# measure signal to noise ratio
signal_2_noise(lbh_selec_table, mar = 0.4)

# measure dynamic time warping
warbleR::df_DTW(lbh_selec_table)

# measure gap duration
warbleR::gaps(lbh_selec_table)

```

Testing coordinated singing:
```{r}
#load  simulated singing data (see data documentation)
data(sim_coor_sing)

# testing if coordination happens less than expected by chance
warbleR::test_coordination(sim_coor_sing)

sim_coor_sing$sound.files <- sim_coor_sing$sing.event
sim_coor_sing$selec <- 1:nrow(sim_coor_sing)

ovlps <- warbleR::overlapping_sels(sim_coor_sing, pb = FALSE)

```


'Song' level features:
```{r}

pe_st <- read.csv("./data/P. eurynome selection table.csv")

p.eurynome <- readWave("./data/Phaethornis-eurynome-15607.wav", from = 0.7, to = 4, units = "seconds")

# check song
seewave::spectro(p.eurynome, flim = c(1, 14), collevels = seq(-120, 0, 5), palette = viridis, scale = FALSE, grid = FALSE, ovlp = 90)

# check elements
seewave::spectro(p.eurynome, flim = c(1, 14),  tlim = c(0, 0.6), collevels = seq(-120, 0, 5), palette = viridis, scale = FALSE, grid = FALSE, ovlp = 90)

# create selection table
pe_est <- warbleR::selection_table(pe_st, extended = TRUE, 
              confirm.extended = FALSE, by.song = "song")

# extract song level features
song_feat <- warbleR::song_param(pe_est, song_colm = "song")

```

other features:
```{r}

check_sels(lbh_selec_table)

library(NatureSounds)
library(viridis)
data("lbh.est")

spectrograms(lbh.est, path = wd, pal = viridis, collevels = seq(-100, 0 , 5), ovlp = 90, wl = 100)

spectrograms(lbh.est, path = wd, pal = viridis, fast.spec = TRUE, wl = 100)


catalog(X = lbh.est, flim = c(1, 10), nrow = 8, ncol = 7, ovlp = 70, height = 15, width = 23, same.time.scale = TRUE, mar = 0.005, wl = 200, gr = FALSE, spec.mar = 0.4, lab.mar = 0.8, rm.axes = TRUE, by.row = TRUE, box = TRUE, labels = "selec", fast.spec = FALSE, pal = viridis, path = wd, collevels = seq(-100, 0 , 5))

```


# quantify acoustic (latent) space: PhenotypeSpace

```{r}

library(PhenotypeSpace)
data("example_space")

# plot data
xs <- tapply(example_space$Dimension_1, example_space$ID, mean)
ys <- tapply(example_space$Dimension_2, example_space$ID, mean)
plot(example_space[, c("Dimension_1", "Dimension_2")], col = example_space$color, pch = 20, cex = 1.8)
text(xs, ys, labels = names(xs), cex = 2.5)


# get minimum convex polygon overlap for each group (non-symmetric)
mcp_overlaps <- space_similarity(
X = example_space,
dimensions =  c("Dimension_1", "Dimension_2"),
group = "ID",
type = "density.overlap")

# convert to non-symmetric triangular matrix
rectangular_to_triangular(mcp_overlaps, symmetric = FALSE)

```

Measure latent space on acoustic data:

```{r}

# measure acoustic parameters
sp <- spectro_analysis(X = lbh.est, bp = c(0, 11), wl = 100, threshold = 1, ovlp = 70)

# reduce dimensions with PCA
pca <- princomp(sp[,-c(1:2)])$scores[, 1:2]

# create data frame
sp <- data.frame(lek.song.type = lbh.est$lek.song.type, pca)

# measure distances between points
ovlp <- space_similarity(X = sp, dimensions = c("Comp.1", "Comp.2"), group = "lek.song.type", type = "distance")

# convert to non-symmetric triangular matrix
rectangular_to_triangular(ovlp, symmetric = TRUE)

```

