---
title: "cito"
author: "Maximilian Pichler"
date: "2026-02-02"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation
```{r}
# devtools::install_github('citoverse/cito') # development version
# devtools::install_github(repo = "TheoreticalEcology/EcoData", dependencies = F, build_vignettes = F) # additional data
library(EcoData) # for elefant data
library(cito)
library(torch)
library(ggplot2)
set.seed(42)
```

## Fully-connected Neural Networks

### Classification and Regression tasks
Loss function, which "likelihood" to use

Regression - continuous response variables
```{r}
nn.fit = dnn(Sepal.Width~Sepal.Length+Petal.Length, data = iris, loss = "mse", batchsize = 30L) # Mean squared error

nn.fit = dnn(Sepal.Width~Sepal.Length+Petal.Length, data = iris, loss = "mse", batchsize = 30L, device = "mps") # change device mps for mac apple silicon and cuda for NVIDIA GPUs

predictions = predict(nn.fit)
plot(iris$Sepal.Width, predictions)
abline(c(0, 1))

print(nn.fit)
coef(nn.fit)
```

```{r}
saveRDS(nn.fit, file = "mymodel.RDS")
nn.fit = readRDS(file = "mymodel.RDS")
```


- Epochs == iterations, DNN are trained via batch stochastic gradient descent. With n = 150 and bs = 30L, one epoch consists of 5 optimization steps!
- lr = learning rate, stepsize of the gradient optimizer
- Plot on the right side, loss during training, should go down!


```{r}
plot(nn.fit)
```

Classification - binary or multiclass
```{r}
nn.fit = dnn(Species~Sepal.Width+Petal.Length, data = iris, loss = "cross-entropy") 
predict(nn.fit) # predict on the scale of the link
predict(nn.fit, type = "response") # predict on the scale of the response
```

```{r}
pred = predict(nn.fit, type = "response") # probabilities that sum up to 1 over the rows
mean(1*(apply(pred, 1, which.max) == as.integer(iris$Species)))
```

### Architecture

Regression - continuous response variables
```{r}
nn.fit = dnn(Sepal.Width~Sepal.Length+Petal.Length, data = iris, loss = "mse", hidden = c(5, 5)) # Mean squared error
plot(nn.fit)
```

```{r}
nn.fit = dnn(Sepal.Width~Sepal.Length+Petal.Length, data = iris, loss = "mse", hidden = c(50, 50)) # Mean squared error
plot(nn.fit)
```

- `hidden` = architecture, each element corresponds to one hidden layer
- Rule of thumb? It is hard to say, we need a specific size so that our model can approximate any function...
- My suggestion is to keep it to 2-3 hidden layers with 50 to 200 neurons


### Convergence 
#### Learning Rate

Problem: First-order gradient descent, high sensitivity to learning rate (=step size of the optimizer)!
  
If learning rate is too low or too high, optimizer does not find into a good local minima

Baseline loss == Intercept only model, the model should get at least better than the baseline loss!

```{r}
set.seed(42)
nn.fit = dnn(Sepal.Width~Sepal.Length+Petal.Length, data = iris, loss = "mse", lr = 0.0000001) # Mean squared error
```

Learning rate is an important hyperparameter!


#### Scaling
```{r}
nn.fit = dnn(Sepal.Width~Sepal.Length+Petal.Length, data = iris, loss = "mse", lr = 0.0000001) # Mean squared error
```


### Overfitting

```{r}
set.seed(42)
# SDM - Predicting Elephant occurrences
elephant = EcoData::elephant
indices = sample.int(nrow(elephant$occurenceData), 500)
df = elephant$occurenceData[indices, ]
df$Presence = as.integer(df$Presence)
head(df)
```
Overfitting --> our model overfits to the data, hence generalize poorly to new data. 

```{r}
nn.fit = dnn(Presence~., 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 500L)
```

Validation data can be used to monitor/detect overfitting:
  
```{r}
nn.fit = dnn(Presence~., 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 500L,
             validation = 0.3)
```

Regularization and/or early stopping prevent overfitting:

- Regularization: l1 (LASSO) and l2 (Ridge) (lambda and alpha parameters in cito), dropout 

```{r}
nn.fit = dnn(Presence~., 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 500L,
             lambda = 0.001, # regularization strength
             alpha = 0.8, # weighting between l1 and l2 (80% l1)
             validation = 0.3)
coef(nn.fit)[[1]][[3]] |> fields::image.plot()
```


```{r}
nn.fit = dnn(Presence~., 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 500L,
             dropout = 0.3,
             validation = 0.3)
```

Both approaches are sota, usually we use a combination of the two. 

- Early stopping: stop training when validation loss starts to increase

```{r}
nn.fit = dnn(Presence~., 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 500L,
             lambda = 0.001, # regularization strength
             alpha = 0.2, # weighting between l1 and l2 (20% l1)
             validation = 0.3, 
             early_stopping = 10L) # number of epochs to wait until training is aborted
```

### Hyperparameter tuning
Hyperparameters (learning rate, regularization parameters) should be tuned under Cross-Validation and not manually. 

cito will tune hyperparameters that are flagged with `=tune()` under cross-validation, will then select the best set of hyperparameters and return a final model that was trained on that best set of hyperparameters


```{r}
nn.fit = dnn(Presence~., 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 500L,
             lambda = tune(),
             alpha = 0.2, # 
             validation = 0.3,
             early_stopping = 10L) # number of epochs to wait until training is aborted
```

`tune` is a function, custom values or possible ranges can be passed 

Tuning history:

```{r}
nn.fit$tuning
```

Tuning setup (how many hyperparameter samples, number of splits in the cross-validation, parallelization, etc.) are set in `tuning = config_tuning(...)`

```{r}
nn.fit = dnn(Presence~., 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 500L,
             lambda = tune(),
             alpha = 0.2, # 
             validation = 0.3,
             early_stopping = 10L,
             tuning = config_tuning(CV = 3L, 
                                    steps = 5L) # number of random samples from the hyperparameter space, here 5 random lambda values
) 
```

Architecture can be also tuned but we discourage from doing that, the hyperparameter space is just too complex/large and there are interactions with other hyperparameters such as regularization or learning rate. We suggest to tune mostly lambda, alpha, and the learning rate!
  
### Inference / Explaining 

Two reasons:
- We want to infer the predictive patterns in the data
- Reliability of the model, the right predictions because of the right reasons
  
  
```{r}
nn.fit = dnn(Presence~bio1+bio2+bio15, 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 50L)

m2 = glm(Presence~bio1+bio2+bio15, data = df, family = "binomial")
```

`summary` will calculate:
  
- Feature importance == permutation importance (similar to the variable importance in random forest)
- average conditional effects (ACE) == approximation of linear effects
- (std of ACE == non-linearity of the effects (will be removed))

```{r}
summary(nn.fit, type = "link")
coef(m2)
car::Anova(m2)
```

ACE are not perfect for non-linear effects, accumulated local effect plots (and partial dependency plots) are typically used for non-linear relationships:
  
```{r}
ALE(nn.fit)
plot(effects::allEffects(m2))
```

standard errors / p-values can be generated via bootstrapping (expensive!):

```{r}
nn.fit = dnn(Presence~bio1+bio2+bio15, 
             data = df, 
             loss = "bernoulli", 
             lr = 0.001, 
             hidden = c(200L, 200L), 
             optimizer = "ignite_adam",
             epochs = 50L,
             bootstrap = 20L, 
             bootstrap_parallel = 5L)
```

```{r}
summary(nn.fit)
```

```{r}
ALE(nn.fit)
```

